Mining HIV Trends in Social Media Data

### 2016-05-03 draft/draft.tex
add references and clean draft

%\usepackage [english]{babel}
%\usepackage [autostyle, english = american]{csquotes}
%\MakeOuterQuote{"}

\begin{document}

% Copyright
\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

@@

Pre-Exposure Prophylaxis (PrEP) is a recently developed therapy for the prevention of Human Immunodeficiency Virus (HIV) transmission. The treatment is available as a pill named Truvada and is taken once a day by HIV negative individuals for as much as 99\% protection when exposed to HIV. While medically effective, PrEP suffers from high drug costs, HIV-related social stigma, and under-informed health providers and patients. Data mining of social media has proven effective in the monitoring of HIV in the US, but no study has investigated PrEP using social media. This paper describes a data mining and machine learning strategy using natural language processing that monitors Twitter social media data to identify PrEP discussion trends. Our results show that we can identify PrEP and HIV discussion dynamics over time, and PrEP-related tweets with positive and negative sentiment. These results can be used by public health professionals to monitor PrEP discussion trends, and identify strategies to improve HIV prevention.

@@

\section{Introduction}
Pre-exposure prophylaxis (PrEP) is an HIV prevention method marketed in the form of a pill named Truvada. Truvada was approved by the Food and Drug Administration (FDA) in 2012 to prevent sexually acquired HIV infection after several positive trials~\cite{grant2010preexposure,thigpen2012antiretroviral} showed that it was safe and effective. Despite minimal side effects, and risk protection of up to 99\%, PrEP suffers from uncertain health insurance compensation, the risk of spreading HIV drug resistance and uninformed health providers and patients~\cite{liu2014early}. Truvada must be taken once a day for full protection. In some individuals, adherence is difficult to maintain~\cite{van2012unraveling}, leading to loss of Truvada's effectiveness. While public health officials can spread information and monitor the effectiveness of PrEP at the clinic-level, scaled-up data mining on social media data may provide more complete information on PrEP at the national level.

Twitter has been used as a source of data for large scale opinion mining in public health monitoring contexts to predict the spread of influenza~\cite{aramaki2011twitter}, predict postpartum depression~\cite{de2013predicting}, and examine tobacco use~\cite{myslin2013using}. Recently it has also been used for the study of HIV~\cite{young2014methods,young2013online}. These studies of HIV have focused on county-level HIV prevalence prediction, and general HIV discussion monitoring, but they have not focused on PrEP related discussion. Furthermore, existing HIV social media analyses have not taken full advantage of natural language processing (NLP) techniques to discover semantic information in unstructured text~\cite{young2015big}.

Tweets from twitter consist of short 140 character text messages that may also contain hashtag annotations.

@@

Latent Dirichlet Allocation~\cite{blei2003latent} (LDA) is used to identify a small set of latent topics present in an unstructured corpus. LDA is a graphical model that generates documents from a set of latent topics. A topic is a probability distribution over words that captures a set of related words. LDA models are often inferred in practice using Bayesian inference via either collapsed Gibbs sampling or Variational Bayesian inference. Inspection of the resulting topics allows one to identify relevant terms and the context in which they occur in the corpus.

Dynamic Topic Modeling~\cite{blei2006dynamic} (DTM) is an extension of LDA that produces a seried of topic models over time. Briefly, the documents in the corpus are divided into several corpuses that are successive in time. LDA is performed on each corpus, to extract a topic distribution. The posterior topic distribution from time $t_n$ is used as the prior for time $t_{n+1}$. This lets the DTM model determine a topic model for each time point that is dependent both on the set of tweets from that time point, and on the previous time point's document distribution.

In this paper we answer the acknowledged need to harness large scale data in the battle against HIV~\cite{young2015big}. We use the above NLP techniques to extract PrEP-related semantic information from a Twitter corpus dataset. We identify critical PrEP related terms, users, hashtags and tweets. We identify PrEP discussion trends over time, and identify other topics that people who tweet about PrEP also tweet about. Finally we train a sentiment classifier that automatically identifies PrEP related tweets with positive and negative sentiment. Together these results, and these approaches can be used by public health officials to identify the national PrEP discussion and respond to issues as they arise.

\section{Results}
We sought to determine trends in HIV and PrEP discourse on twitter to inform and coordinate public health efforts aiming to promote PrEP adoption and adherence for at-risk individuals. We collected 624,569 tweets containing at least one of the following words 'HIV', 'AIDS', 'truvada', 'prophylaxis', 'imtesting', 'PrEP' from Twitter's streaming API (TODO: how to cite Twitter streaming API?). The tweets were restricted to English language, and the collection dates spanned from the 47th week of 2015 to the 14th week of 2016. The tweets were cleaned of exotic characters. Before performing topic analysis, we excluded words that were mentioned fewer than 10 times, or more than 0.3 times the number of documents. We also performed Term Frequency Inverse Document Frequency (TF-IDF) normalization.

We excluded tweets that did not originate in contiguous United States time zones. 14204 of the tweets we collected (about 2\%) had geolocation coordinates available. We found that tweets were largely concentrated on US and Canadian metro areas (Figure 1). The tweets did not seem to be over represented in any geographical region of the US.

@@

We trained a Word2Vec model, and queried for the top 10 word-vectors related to the term PrEP (Table 1). We found several HIV/AIDS related events, WorldAIDSDay, HLM2016AIDS, ICASA2015, as well as the PrEP drug Truvada, and the term HIV. In addition, we found the acronym ART which refers to Anti-Retroviral Therapy. DoingIt, and OneConversation, are ongoing efforts led by the Center for Disease Control (CDC) to spread awareness and reduce the spread of HIV. NBHAAD is an organization that is committed to increasing awareness for HIV within the Black community. Nancy Reagan, who died in early 2016, was mentioned in conjunction with her efforts to combat HIV in the 1980's. Together these results show us a high-level view of the important components of the national PrEP conversation on Twitter.

@@

Interestingly, we also found 3 tweets and 1 user in the top 10 doc2vec results for \#prep. Tweet 702179860983189504 has content spreading HIV prevention awareness: "\#StoneColdVideoTODAY if You see this 13 symptoms. Do HIV Test Immediately. Must Read". Conversely tweet 708519265540907010 has content that calls into doubt the usefulness of PrEP: "Checkout why PrEP is hurting the cause \& \#JoinTheConversation \#LGBTQIA". User 711275699529764864 appears to be a twitter spam bot with no obvious connection to PrEP.  Thus with this method we demonstrate a way to identify and monitor the most viral tweets related to PrEP. As the discussion changes, public health professionals can use this information to quickly identify the most relevant viral sentiment in the online PrEP conversation.

% table 2 - d2v PrEP
\begin{table}
\centering
\caption{Cosine Similarity to doc-vector \#PrEP}
\begin{tabular}{|l|c|} \hline
Related Hashtag/Tweet & Cosine Similarity to \#PrEP\\ \hline
\#lgbtmedia16 & 0.739128\\ \hline
\#hiv & 	0.727602 \\ \hline
\#whereisprep & 0.707165 \\ \hline
\#truvada & 0.696113 \\ \hline
\#hivprevention & 0.636068 \\ \hline
tweet-702179860983189504 & 0.630055\\ \hline
user-711275699529764864 & 0.629254\\ \hline
tweet-708519265540907010 & 0.628778 \\ \hline
tweet-712032637024653313 & 0.628646 \\ \hline
\#harrogatehour & 0.628547 \\ \hline
\hline\end{tabular}
\end{table}

Finally we wanted to visualize the relative similarities of several PrEP-related keywords in a low dimensional space. We took keywords that we had identified in our PrEP related queries, along with other HIV-prevention related terms, and visualised their word-vectors in 2 dimensions using tSNE~\cite{van2008visualizing} (Figure 2.

We several trends. Notably the pharmaceutical based HIV therapies all cluster together (ART, PrEP, truvada) and the AIDS awareness events cluster together (WorldAIDSDay, NBHAAD, ICASA2015). Words that are related to HIV discussion, but also used in other contexts (undetectable, testing, awareness) are further away from the HIV/AIDS word-clusters. These results provide another mechanism for researchers to visualize and identify relevant trends in PrEP-related keywords.

@@

We found at least one other DTM topic that showed interesting behaviour. We found that topic 4 captured several keywords related to World AIDS Day (Figure 4). We can see that "WorldAIDSDay", and "Can" peak in the 47th week of 2015 and then decline into 2016. This correlates well with the actual date of World AIDS Day, December 1st. Furthermore, while December 1st is World AIDS day, the whole month of December is AIDS Awareness Month. We can clearly see the words 'raise' and 'awareness' peak later and last longer than the word 'WorldAIDSDay' implying that it is correlated to the whole month of December. While our PrEP investigation isn't specifically interested in World AIDS Day, or AIDS Awareness Month, this observation validates our ability to accurately identify temporal events using DTM.

@@

We wanted to identify what Twitter users that mentioned PrEP were discussing in their other tweets. We identified users that were most similar to PrEP according to our Paragraph2Vec results, and downloaded their most recent 3000 tweets. We took the top 500 users that had at least 200 words in the combined tweets of their tweet history. For each user, we concatenated all of their tweets, and performed LDA topic modelling on the resulting set of user-timeline documents (Figure 5).

@@

We performed LDA on the users' timelines using the pyLDAvis python package from graphlab~\cite{low2014graphlab}. We specified 100 topics, and found that only one of them was related to HIV (topic number13, see Figure 5) representing about 2\% of the marginal topic distribution. Topic 13 contained top terms "hivaids", "pepfar", "stigma" and "aidsgov". PEPFAR is a governmental organization, The United States President's Emergency Plan for AIDS Relief, that is focused on combating the spread of AIDS internationally. Interestingly, we don't see PrEP in the top 30 terms in topic 13. This indicates that among twitter users that talk about PrEP, HIV is a small part of their discussion, and PrEP an even smaller part of their discussion.

@@

We chose to use a relatively simple classifier model (logistic regression) and stop training at 20 iterations of stochastic gradient decent over the corpus, because we wanted to prevent the possibility that we overtrained on our training data. This was especially important because our training and testing data, while both sets of tweets, were separate datasets. We then used this classifier to classify our PrEP related tweets into positive or negative sentiment labels. We identified the most positive, and most negative tweets, by log probability, on our full dataset, and on tweets that specifically mention either PrEP or Truvada. We provided the text from the top 3 positive and negative tweets from each dataset (Table 3, Table 4).

@@

While the sentiment classification doesn't have perfect accuracy, we can see positive tweets disseminating information about the efficacy and benefits of PrEP related preventions (Table 3). The positive tweets indicate that PrEP public health efforts have had some effect disseminating PrEP information on Twitter.

The negative tweets may be more important than the positive tweets to guide future public health policy corrections. The negative tweets contain concerns that Truvada may not block HIV transmission in all cases (Table 4). There also seems to be concern that use of Truvada may increase the transmission of non-HIV STDs such as Hepatitis C because some PrEP users stop using condoms. This may indicate that public health professionals need to stress that PrEP users should not stop wearing condoms when they spread PrEP related information.

Another negative tweet contains questions over weather Truvada is available as a generic drug. This may indicate that in some cases, patients are aware of PrEP, but do not have access to it, either because it is not available through their healthcare provider, or because it is not affordable.

@@

>>>>

%\usepackage [english]{babel}
%\usepackage [autostyle, english = american]{csquotes}
%\MakeOuterQuote{"}
\usepackage{lineno}
\begin{document}
\linenumbers
% Copyright
\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

@@

Pre-Exposure Prophylaxis (PrEP) is a recently developed therapy for the prevention of Human Immunodeficiency Virus (HIV) transmission. The treatment is available as a pill named Truvada and is taken once a day by HIV negative individuals for as much as 99\% protection when exposed to HIV. While medically effective, PrEP suffers from high drug costs, HIV-related social stigma, and under-informed health providers and patients. Data mining of social media has proven effective in the monitoring of HIV in the United States, but no study has investigated PrEP using social media. This paper describes a data mining and machine learning strategy using natural language processing (NLP) that monitors Twitter social media data to identify PrEP discussion trends. Our results show that we can identify PrEP and HIV discussion dynamics over time, and PrEP-related tweets with positive and negative sentiment. These results can be used by public health professionals to monitor PrEP discussion trends, and identify strategies to improve HIV prevention.

@@

\section{Introduction}
Pre-exposure prophylaxis (PrEP), a treatment for the prevention of Immunodeficiency Virus (HIV), is marketed under the trade name Truvada. Truvada was approved by the Food and Drug Administration (FDA) in 2012 to prevent sexually acquired HIV infection after several positive trials~\cite{grant2010preexposure,thigpen2012antiretroviral} showed that it was safe and effective. Despite minimal side effects, and risk protection of up to 99\%, PrEP suffers from uncertain health insurance compensation, the risk of producing drug resistant strains and uninformed health providers and patients~\cite{liu2014early}. Truvada must be taken once a day for full protection. In some individuals, adherence is difficult to maintain~\cite{van2012unraveling}, leading to loss of Truvada's effectiveness. While public health officials can spread information and monitor the effectiveness of PrEP at the clinic-level, scaled-up data mining on social media data may provide more complete information on PrEP at the national level.

Tweets from twitter consist of short 140 character text messages that may also contain hashtag annotations. Twitter has been used as a source of data for large scale opinion mining in public health monitoring contexts to predict the spread of influenza~\cite{aramaki2011twitter}, predict postpartum depression~\cite{de2013predicting}, and examine tobacco use~\cite{myslin2013using}. Recently it has also been used for the study of HIV~\cite{young2014methods,young2013online}. These studies of HIV have focused on county-level HIV prevalence prediction, and general HIV discussion monitoring, but they have not focused on PrEP related discussion. Furthermore, existing HIV social media analyses have not taken full advantage of natural language processing (NLP) techniques to discover semantic information in unstructured text~\cite{young2015big}.

@@

Latent Dirichlet Allocation~\cite{blei2003latent} (LDA) is used to identify a small set of latent topics present in an unstructured corpus. LDA is a graphical model that generates documents from a set of latent topics. A topic is a probability distribution over words that captures a set of related words. LDA models are often inferred in practice using Bayesian inference either via collapsed Gibbs sampling or Variational Bayesian inference. Inspection of the resulting topics allows one to identify relevant terms and the context in which they occur in the corpus.

Dynamic Topic Modelling~\cite{blei2006dynamic} (DTM) is an extension of LDA that produces a seried of topic models over time. Briefly, the documents in the corpus are divided into several corpuses that are successive in time. LDA is performed on each corpus to extract a topic distribution. The posterior topic distribution from time $t_n$ is used as the prior for time $t_{n+1}$. This lets the DTM model determine a topic model for each time point that is dependent both on the set of tweets from that time point, and on the previous time point's topic distribution.

In this paper we respond to the acknowledged need to harness large scale data in the battle against HIV~\cite{young2015big}. We use the above NLP techniques to extract PrEP-related semantic information from a Twitter corpus dataset. We identify critical PrEP related terms, users, hashtags and tweets. We identify PrEP discussion trends over time, and identify other topics that people who tweet about PrEP, also tweet about. Finally we train a sentiment classifier that automatically identifies PrEP-related tweets with positive and negative sentiment. Together these results, and these approaches, can be used by public health officials to identify the national PrEP discussion and respond to issues as they arise.

\section{Results}
We sought to determine trends in HIV and PrEP discourse on twitter to inform and coordinate public health efforts aiming to promote PrEP adoption and adherence for at-risk individuals. We collected 624,569 tweets containing at least one of the following words 'HIV', 'AIDS', 'truvada', 'prophylaxis', 'imtesting', 'PrEP' from Twitter's streaming API. The tweets were restricted to English language, and the collection dates spanned from the 47th week of 2015 to the 14th week of 2016. The tweets were cleaned of exotic characters. Before performing topic analysis, we excluded words that were mentioned fewer than 10 times, or more than 0.3 times the number of documents. We also performed Term Frequency Inverse Document Frequency (TF-IDF) normalization.

We excluded tweets that did not originate in contiguous United States time zones. A subset of 14,204 tweets (about 2\%) had geolocation coordinates available. We found that these tweets were largely concentrated on US and Canadian metro areas (Figure 1). The tweets did not seem to be over represented in any geographical region of the US.

@@

We trained a Word2Vec model, and queried for the top 10 word-vectors related to the term PrEP (Table 1). We found several HIV/AIDS related events, WorldAIDSDay, HLM2016AIDS, ICASA2015, as well as the PrEP drug Truvada, and the term HIV. In addition, we found the acronym ART which refers to Anti-Retroviral Therapy. DoingIt, and OneConversation, are ongoing efforts led by the Center for Disease Control (CDC) to spread awareness and reduce the spread of HIV. NBHAAD is an organization that is committed to increasing awareness for HIV within the Black community. Nancy Reagan, who died in early 2016, was mentioned in conjunction with her efforts to combat HIV in the 1980's. Together these results show us a high-level view of the important components of the national PrEP conversation on Twitter over the collection period.

@@

Interestingly, we also found 3 tweets and 1 user in the top 10 doc2vec results for \#prep. Tweet 702179860983189504 has content spreading HIV prevention awareness: "\#StoneColdVideoTODAY if You see this 13 symptoms. Do HIV Test Immediately. Must Read". Conversely tweet 708519265540907010 has content that calls into doubt the usefulness of PrEP: "Checkout why PrEP is hurting the cause \& \#JoinTheConversation \#LGBTQIA".

TODO: investigate those tweets, especially "Checkout why PrEP is hurting the cause"

User 711275699529764864 appears to be a twitter spam bot with no obvious connection to PrEP. Thus while we observe some noise, our results demonstrate a way to quickly identify and monitor the most viral tweets related to PrEP. As the discussion changes, public health professionals can use this information to quickly identify the most relevant viral sentiment in the online PrEP conversation.

% table 2 - d2v PrEP
\begin{table}
\centering
\caption{Cosine Similarity to doc-vector \#PrEP}
\begin{tabular}{|l|c|} \hline
Related hashtag/Tweet & Cosine Similarity to \#PrEP\\ \hline
\#lgbtmedia16 & 0.739128\\ \hline
\#hiv & 	0.727602 \\ \hline
\#whereisprep & 0.707165 \\ \hline
\#truvada & 0.696113 \\ \hline
\#hivprevention & 0.636068 \\ \hline
tweet-702179860983189504 & 0.630055\\ \hline
user-711275699529764864 & 0.629254\\ \hline
tweet-708519265540907010 & 0.628778 \\ \hline
tweet-712032637024653313 & 0.628646 \\ \hline
\#harrogatehour & 0.628547 \\ \hline
\hline\end{tabular}
\end{table}

Finally we wanted to visualize the relative similarities of several PrEP-related keywords in a low dimensional space. We took keywords that we had identified in our PrEP related queries, along with other HIV-prevention related terms, and visualised their word-vectors in 2 dimensions using t-distributed Stochastic Neighbour Embedding(tSNE)~\cite{van2008visualizing} (Figure 2.

We identified several trends. Notably the pharmaceutical based HIV therapies all cluster together (ART, PrEP, truvada) and the AIDS awareness events cluster together (WorldAIDSDay, NBHAAD, ICASA2015). Words that are related to HIV discussion, but also used in other contexts (undetectable, testing, awareness) are further away from the HIV/AIDS word-clusters. These results provide another mechanism for researchers to visualize and identify relevant trends in PrEP-related keywords.

@@

We found at least one other DTM topic that showed interesting behaviour. We observed that topic 4 captured several keywords related to World AIDS Day (Figure 4). We can see that "WorldAIDSDay", and "Can" peak in the 47th week of 2015 and then decline into 2016. This correlates well with the actual date of World AIDS Day, December 1st. Furthermore, while December 1st is World AIDS day, the whole month of December is AIDS Awareness Month. We can clearly see the words 'raise' and 'awareness' peak later and last longer than the word 'WorldAIDSDay' implying that it is correlated to the whole month of December. While our PrEP investigation isn't specifically interested in World AIDS Day, or AIDS Awareness Month, this observation serves as a control to validate our ability to accurately identify temporal events using DTM.

@@

We wanted to identify what Twitter users that mentioned PrEP were discussing in their other tweets. We identified users that were most similar to PrEP according to our Paragraph2Vec results, and downloaded their recent tweet history, up to their last 3000 tweets. We selected top 500 users that had at least 200 words in the combined tweets of their tweet history. For each user, we concatenated all of their tweets, and performed LDA topic modelling on the resulting set of user-timeline documents (Figure 5).

@@

We performed LDA on the users' timelines using the pyLDAvis python package from graphlab~\cite{low2014graphlab}. We specified 100 topics, and found that only one of them was related to HIV (topic number 13, see Figure 5) representing about 2\% of the marginal topic distribution. Topic 13 contained top terms "hivaids", "pepfar", "stigma" and "aidsgov". PEPFAR is a governmental organization, The United States President's Emergency Plan for AIDS Relief, that is focused on combating the spread of AIDS internationally. Interestingly, we don't see PrEP in the top 30 terms in topic 13. This indicates that among twitter users that talk about PrEP, HIV is a small part of their discussion, and PrEP an even smaller part of their discussion.

@@

We chose to use a relatively simple classifier model (logistic regression) and stop training at 20 epochs of stochastic gradient decent over the corpus, because we wanted to prevent the possibility that we overtrained on our training data. This was especially important because our training and testing data, while both sets of tweets, were separate datasets. We then used this trained classifier to classify our PrEP related tweets into positive or negative sentiment labels. We identified the most positive, and most negative tweets, by log probability, on our full dataset, and on tweets that specifically mention either PrEP or Truvada. We provide the text from the top three positive and negative tweets from each of the three datasets (Table 3, Table 4).

@@

While the sentiment classification doesn't have perfect sentiment accuracy, we can see in general that the positive tweets are disseminating information about the efficacy and benefits of PrEP related preventions (Table 3). The positive tweets indicate that PrEP public health informational efforts have had some effect disseminating PrEP information on Twitter.

The negative tweets may be more important than the positive tweets to guide future public health policy corrections. The negative tweets contain concerns that Truvada may not block HIV transmission in all cases (Table 4). There also seems to be concern that use of Truvada may increase the transmission of non-HIV STDs such as Hepatitis C because some PrEP users stop using condoms. This may indicate that public health professionals need to stress that PrEP users continue to wear condoms for full protection against other diseases in PrEP-related health information.

Another negative tweet questions whether Truvada is available as a generic drug. PrEP is currently only available through Gilead's Truvada, though a generic may be available as soon as 2017~\cite{truvadagenericblog}. Current patients must deal with uncertain drug prices, which vary from \$14,000 to \$70~\cite{truvadagenericblog}, and have in some cases, uncertain medical insurance reimbursement status. This tweet highlights both the difficulty of aquiring affordable PrEP medication, and also a level of conflicting information circulating on social media.

@@

TODO: investigate the reference to  \#truvadawhore which came up in the sentiment tweets above.
